---
title: "Data Cleaning"
author: "Jack A. Goldman"
date: "2025-12-26"
format: html
---


```{r}
# Load necessary libraries
library(dplyr)
library(tidyverse)
library(readr)
library(zoo)

# load fire information
fire_info <- read_csv("data/fire_information.csv") |> 
        filter(!coalesce(PRESCRIBED, FALSE))  # remove fires where PRESCRIBED == TRUE
```


# Purpose 

The purpose of this script is to clean and preprocess the raw dataset for further analysis. 

# VPD data cleaning for lagged analysis
```{r}
# Load the raw dataset
vpd_data <- read_csv("data/vpd_weather_daily_processed_v2.csv")
# Inspect the data
head(vpd_data)
str(vpd_data)
```


For each GID filter date so that we only have dates within 30 days before fire start date (fire starte date is the Start_Date column in fire_info) and add column lag that is 1-30 for days before fire start date

```{r}
vpd_30 = vpd_data |> 
    left_join(fire_info |> select(GID, Start_Date), by = "GID") |> 
    mutate(
        date = as.Date(date),
        Start_Date = as.Date(Start_Date)
    ) |> 
    filter(date >= (Start_Date - 30) & date < Start_Date) |>  # keep 1–30 days before start
    mutate(lag = as.integer(Start_Date - date)) |>            # lag = 1..30
    select(-Start_Date)

```


VPD matrix for distributed lag model
Currently in long format, need to convert to wide format with each column as vpd for a specific lag and each row as a fire event (GID) but arrange columns by lag vpd_lag_1 to vpd_lag_30
```{r}
vpd_wide = vpd_30 %>%
    select(GID, lag, vpd) %>%
    arrange(lag) %>%
    pivot_wider(names_from = lag, values_from = vpd, names_prefix = "vpd_") %>%
    drop_na() |> 
    arrange(GID)

vpd_wide_gids = unique(vpd_wide$GID)
nrow(vpd_wide)
```

Save vpd_wide as vpd_30_wide.csv
```{r}
saveRDS(vpd_wide, "data/cleaned/vpd_30_wide.rds")
```

# vpd data cleaning for moving average over 30 days
```{r}

vpd_ma_30 = vpd_data |> 
    left_join(fire_info |> select(GID, Start_Date), by = "GID") |> 
    mutate(
        date = as.Date(date),
        Start_Date = as.Date(Start_Date)
    ) |> 
    arrange(GID, date) |> 
    group_by(GID) |> 
    filter(date >= (Start_Date - 30) & date < Start_Date) |> 
    mutate(vpd_ma_30 = zoo::rollapply(vpd, width = 30, FUN = function(x) mean(x, na.rm = TRUE), fill = NA, align = "right")) |> 
    summarize(
        vpd_ma_30 = { tmp <- na.omit(vpd_ma_30); if (length(tmp)) tmp[length(tmp)] else NA_real_ },
        .groups = "drop"
    ) |> 
    drop_na() 
```

Save vpd_ma_30 

```{r}
saveRDS(vpd_ma_30, "data/cleaned/vpd_ma_30.rds")
```

 # RBR data cleaning
 Load RBR zonal stats data and select only GID, median, and 90th percentile

 ```{r}
rbr_data = read_csv("data/batch_rbr_w_offset_zonal_stats.csv") |> 
    select(GID, median, percentile_90)  |> 
    rename(pct_90 = percentile_90) 

 ```

 Save cleaned RBR data
 ```{r}
saveRDS(rbr_data, "data/cleaned/rbr_cleaned.rds")
 ```

# Landcover cleaning


```{r}
landcover_data = read_csv("data/fire_landcover_pct_v1.csv") 

head(landcover_data)

landcover_cleaned = landcover_data |> 
    select(GID, coniferous, percent_upland, percent_wetland) |> 
    rename(pct_conf = coniferous, 
           pct_up = percent_upland, 
           pct_wet = percent_wetland)

saveRDS(landcover_cleaned, "data/cleaned/landcover_cleaned.rds")
```

# fire information cleaning

```{r}
fire_info_cleaned = fire_info |> 
        select(GID, YEAR, POLY_HA, Start_Date) |> 
        rename(fire_size_ha = POLY_HA, 
               year = YEAR, 
               start_date = Start_Date) |> 
        mutate(month = as.integer(format(as.Date(start_date), "%m")))

```

if fire burned before june 1 set it as spring fire season else summer fire season 
The reason for this is because aspen/deciduous leaf out changes the flammability of the landscape and has been shown to impact fire intensity/severity between spring/summer (Podur and Martell 2009; Parisien et al. 2023)

```{r}
fire_info_cleaned = fire_info_cleaned |> 
    mutate(season = ifelse(month < 6, "spring", "summer"))
```

save the file

```{r}
saveRDS(fire_info_cleaned, "data/cleaned/fire_info_cleaned.rds")
```


# Ecoregion

make sure there are no duplicate GIDs in aou_fire_ecoregion.csv, if there are duplicates, keep the first one and drop the rest
```{r}
ecoregion_data = read_csv("data/aou_fire_ecoregion.csv") |> 
    distinct(GID, .keep_all = TRUE)

saveRDS(ecoregion_data, "data/cleaned/ecoregion_cleaned.rds")
```
